---
title: "dplyr joins tutorial"
author: "Michael Hallquist"
date: "8/15/2017"
output:
  html_document:
    theme: spacelab
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
if (!require(pacman)) { install.packages("pacman"); library(pacman) }
p_load(dplyr, readr)
```
## Introduction to joining datasets using dplyr

This document is a minor adaptation of Jenny Bryan's dplyr joins tutorial: <http://stat545.com/bit001_dplyr-cheatsheet.html>. The goal is to develop an intuition of the four major types of two-table join operations: inner, left, right, and full.

First, we import the data into data.frame objects.
```{r}

superheroes <-"
name, alignment, gender,         publisher
Magneto,       bad,   male,            Marvel
Storm,      good, female,            Marvel
Mystique,       bad, female,            Marvel
Batman,      good,   male,                DC
Joker,       bad,   male,                DC
Catwoman,       bad, female,                DC
Hellboy,      good,   male, Dark Horse Comics
"
superheroes <- read_csv(superheroes, trim_ws = TRUE, skip = 1)

publishers <- "
publisher, yr_founded
DC,       1934
Marvel,       1939
Image,       1992
"
publishers <- read_csv(publishers, trim_ws = TRUE, skip = 1)
```

## Mutating joins: inner, left, right, outer

"Mutating" joins combine variables from two datasets on the basis of one or more keys that match between datasets.

*NB*: By default, *dplyr* will search for common columns across datsets as the matching
keys (natural join). If you want to control the process, specify the key using "by"

###inner join
Require match in both datasets (non-matching rows are dropped)
```{r}
ijsp <- inner_join(superheroes, publishers)
print(ijsp)
```

Same idea, just explicit declaration of key
```{r}
inner_join(superheroes, publishers, by="publisher")
```

###left join
Keep all rows in left-hand 'x' dataset (i.e., superheroes). Add columns from publishers where
there is a match. Fill in NA for non-matching observations.

```{r}
ljsp <- left_join(superheroes, publishers)
print(ljsp)
```

###right join
Keep all rows in right-hand 'y' dataset (i.e., publishers). Add columns from superheroes where
there is a match. Fill in NA for non-matching observations.
```{r}
superheroes %>% right_join(publishers)
```

###full join
Keep all rows in left-hand 'x' (superheroes) and right-hand 'y' (publishers) datasets. 
Resulting dataset will have all columns of both datasets, but filling in
NA for any non-matches on either side.
```{r}
fjsp <- full_join(superheroes, publishers)
print(fjsp)
```

##One-to-many join
Note that when there are non-unique matches, the join adds all possible combinations.
This occurs in a one-to-many join.
```{r}
df1 <- data_frame(x = c(1, 1, 2), y = 1:3)
df2 <- data_frame(x = c(1, 1, 2), z = c("a", "b", "a"))

df1 %>% left_join(df2)
```

##Filtering joins: semi_join and anti_join
Filtering joins use specific criteria to identify observations (rows) from one table that exist or don't exist in another table.

These joins are typically used for diagnosing mismatch between two overlapping datasets.

###semi_join
retain observations (rows) in x that match in y
```{r}
semi_join(x = publishers, y = superheroes)
```

Now, observations in superheroes that match in publishers
```{r}
semi_join(superheroes, publishers)
```

###anti_join
observations in x that are not matched in y
Note that this is similar to setdiff in base R

Observations in superheroes that don't match in publishers
```{r}
anti_join(superheroes, publishers)
```

Now, observations in publishers that don't match in superheroes
```{r}
publishers %>% anti_join(superheroes)
```

##Joining multiple datasets

Fancy approach to multiple dataset merge. Joins datasets two at a time from left to right in the list. The result of a two-table join becomes the 'x' dataset for the next join of a new dataset 'y'.

```{r}
df1 <- data.frame(id=1:10, x=rnorm(10), y=runif(10))
df2 <- data.frame(id=1:11, z=rnorm(11), a=runif(11))
df3 <- data.frame(id=2:10, b=rnorm(9), c=runif(9))

Reduce(function(...) { full_join(...) }, list(df1, df2, df3))
```

Alternative using pipeline (less extensible)
```{r}
mergedf = df1 %>% full_join(df2) %>% full_join(df3)
```