---
title: "bootcamp-survey"
author: "Rick Gilmore"
date: "`r Sys.time()`"
output: 
  github_document:
    toc: true
    toc_depth: 3
  pdf_document:
    toc: true
    toc_depth: 3
  ioslides_presentation:
    widescreen: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Goals

- Download and clean data from 2017 R Bootcamp Survey
- Visualize data
- Prepare reports in `ioslides_presentation`, `pdf_document`, and `word_document` formats

## Preliminaries

Load required packages.

```{r load-packages}
library(tidyverse)
library(googlesheets)
```

## Load data and examine

The survey data are stored in a [Google Sheet](https://docs.google.com/spreadsheets/d/1Ay56u6g4jyEEdlmV2NHxTLBlcjI2gHavta-Ik0kGrpg/edit#gid=896447063). We'll use the `googlesheets` package to open it and create a data frame. Documentation about the package can be found [here](https://cran.r-project.org/web/packages/googlesheets/vignettes/basic-usage.html).

There are some idiosyncrasies in using the `googlesheets` package in an R Markdown document, so I created a separate R script, `get-bootcamp-googlesheet.R` to extract the survey data and save it to a CSV under `data/survey.csv`. We can then just load this file.

---

```{r load-data}
survey <- read_csv("../data/survey.csv")
survey
```

Notice that the `get-bootcamp-googlesheet.R` script changed the names of the variables a bit. We may also want to modify the levels of the `R_exp` variable to make it an ordered factor.

```{r modify-r-exp}
survey$R_exp <- ordered(survey$R_exp, levels=c("No experience",
                                               "Limited experience",
                                               "Extensive",
                                               "I should be teaching this course"))
```

## Visualization and analysis

Now, we can ask important questions.

```{r got-vs-r-exp}
got_vs_r_exp <- survey %>%
  ggplot() +
  aes(x=GoT, y=Age_yrs) +
  facet_grid(. ~ R_exp) +
  geom_point()
got_vs_r_exp
```

